Would NEMO benefit from using the \psyclone\ system?

### 2015-12-02 docs/NEMO/psyclone_and_nemo.tex
NEMO: tweak example text
NEMO: updates to psyclone_and_nemo discussion doc following feedback from Mike Bell

Fundamental to the use of \psyclone\ is the \psykal\ (Parallel System,
Kernel, Algorithm) separation of concerns. In this approach a program
is re-structured into three layers; the Algorithm, PSy and Kernel
layers.  The Algorithm and Kernel layers are the responsibility of the
natural scientist while all code related to parallelism is contained
within the PSy layer. Thus the scientist does not need to concern
themselves with optimisation issues, such as loop fusion, redundant
computation in halos, etc. and parallelism issues, such as placement
of directives and halo swap calls.

@@

The advantage of a recipe is that it can be changed, {\bf with no
  change to the science code}. Thus, for example, one recipe could
target MPI and OpenMP using the Intel compiler on a Cray with a
particular configuration of NEMO. Another recipe could target MPI and
OpenACC using the PGI compiler on GPU's for a different NEMO
configuration. One of the problems with performance portability is
that optimisations for one architecture are not necessarily beneficial
for another. It is difficult enough to get different versions of the
same compiler on the same architecture to work well on a complex code
without changes to source code. If such optimisations are hard coded
then it is very difficult to change the code to work well on another
architecture and almost impossible to have performance portability
with a single code base. \psyclone\ aims to address this problem.

@@

@@

@@

Although the core of NEMO contains only two active tracers
(Temperature and Salinity) it can be used with biogeochemical
models. In those configurations there can be of order $\sim 50$ tracer
species, all of which may be treated independently. This has
significant implications for both data structures and our ability to
exploit the additional parallelism that is available.

>>>>

Fundamental to the use of \psyclone\ is the \psykal\ (Parallel System,
Kernel, Algorithm) separation of concerns. In this approach a program
is re-structured into three layers; the Algorithm, PSy and Kernel
layers.  The Algorithm and Kernel layers are the responsibility of the
natural scientist while all code related to parallelism is contained
within the PSy layer. Thus the scientist does not need to concern
themselves with optimisation issues, such as loop fusion, redundant
computation in halos, etc. and parallelism issues, such as placement
of directives and halo swap calls. This separation also improves the
readability and modularity of code and thus greatly aids with
maintenance.

@@

The advantage of a recipe is that it can be changed, {\bf with no
  change to the science code}. Thus, for example, one recipe could
target MPI and OpenMP using the Intel compiler on a Cray with a
particular configuration of NEMO. Another recipe could target MPI and
OpenACC using the PGI compiler on GPU's for a different NEMO
configuration. One of the problems with performance portability is
that optimisations for one architecture are not necessarily beneficial
for another. It is difficult enough to get different versions of the
same compiler on the same architecture to work well on a complex code
without changes to source code. If such optimisations are hard coded
then it is very difficult to change the code to work well on another
architecture and almost impossible to have performance portability
with a single code base. The situation is even more complex if, as
with NEMO, a single code base supports different model configurations
that may themselves require different optimisations. \psyclone\ aims
to address this problem by keeping the optimisation space fully
separate from the science code.

@@

Note that this example has only been produced to inform this
document. We make no claim that the meta-data etc. is correct or
sufficient for NEMO.

@@

\subsection{Run-time Switches}

In common with most simulation packages, NEMO makes use of run-time
flags in order to enable and disable certain functionality. This
becomes an issue for PSyclone when this functionality relates to
choices in the specific Algorithm that is run. As an example, the
ldf\_slp routine contains code that is only required if partial steps
are in use:

\begin{verbatim}
DO jk = 1, jpk             !==   i- & j-gradient of density   ==!
   DO jj = 1, jpjm1
      DO ji = 1, fs_jpim1   ! vector opt.
         zgru(ji,jj,jk) = umask(ji,jj,jk) * ( ... )
      END DO
   END DO
END DO
IF( ln_zps ) THEN    ! partial steps correction at  bottom ocean level
   DO jj = 1, jpjm1
      DO ji = 1, jpim1
         zgru(ji,jj,mbku(ji,jj)) = ...
      END DO
   END DO
ENDIF
!
zdzr(:,:,1) = 0._wp  !== Local vertical density gradient at T-point == !
DO jk = 2, jpkm1
   zdzr(:,:,jk) = ...
END DO
\end{verbatim}

As PSyclone currently has no support for control-flow statements
within an Invoke, the Algorithm Layer for this example would have to
contain three separate Invokes and might look something like:

\begin{verbatim}
CALL invoke( density_igradient(...), &
             density_jgradient(...) )
IF( ln_zps ) THEN
   CALL invoke( partial_steps_correction(...) )
END IF
CALL invoke( local_density_kgradient(...) )
\end{verbatim}

Since optimisation can only be performed within an Invoke, this
structure might result in less performant code. If this proved to be
the case then one might consider re-writing the Algorithm as:

\begin{verbatim}
IF( ln_zps ) THEN
   CALL invoke( density_igradient(...),        &
                density_jgradient(...),        &
                partial_steps_correction(...), &
                local_density_kgradient(...) )
ELSE
   CALL invoke( density_igradient(...),        &
                density_jgradient(...),        &
                local_density_kgradient(...) )
END IF
\end{verbatim}

This issue is not NEMO specific and in fact is currently under
discussion within the GungHo project. Any solution developed there
will be immediately applicable to NEMO.

@@

Although the core of NEMO contains only two active tracers
(Temperature and Salinity) it can be used with biogeochemical
models. In those configurations there can be of order $\sim 50$ tracer
species. Although these can be treated independently within the
advection routines there are interactions between them in the
biological routines. This has significant implications for both data
structures and our ability to exploit the additional parallelism that
is available. For the latter we would need additional meta-data to
describe this parallelism and, crucially, indicate when it is not
available.
